---
title: "Data Science Specialization - Course of Reproducible Research"
subtitle: "Project Course 1"
author: "David Lira"
date: "16/4/2017"
output: md_document
---
**1. Code for reading in the dataset and/or processing the data**
 
Loading necessary libraries for the project and unzipping the dataset from the cloud using `data.table`.

```{r setup, echo=TRUE, message=FALSE}


library(data.table)
library(dplyr)
library(ggplot2)
library(timeDate)

data <- fread("curl https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip | funzip")
```

**2. Histogram of the total number of steps taken each day**

Creating data.frame with the sum of steps taken daily and corresponding histogram (omitting missing values).

```{r, echo=TRUE, message=FALSE, warning=FALSE}

total_steps = data %>% na.omit %>% group_by(date) %>% summarise(total=sum(steps))


ggplot(total_steps, aes(total)) + geom_histogram(bins=7, col="white", fill="navyblue", alpha=.8)+labs(title="Steps taken each day")

```

**3. Mean and median number of steps taken each day**

Estimators calculated after omitting missing values.

```{r, echo=TRUE, message=FALSE, warning=FALSE}
mean(total_steps$total, na.rm = TRUE)
median(total_steps$total, na.rm = TRUE)

```

**4. Time series plot of the average number of steps taken**

Averaging steps taken across all 5-min intervals and corresponding plot.

```{r, echo=TRUE, message=FALSE, warning=FALSE}

avg_5_min = data %>% na.omit %>% group_by(interval) %>% summarise(avg=mean(steps))

ggplot(avg_5_min, aes(x=interval,y=avg))+geom_line()

```

**5. The 5-minute interval that, on average, contains the maximum number of steps**

According to data, between 8:35 and 8:40 a.m., subject takes more steps on average, around 206.

```{r, echo=TRUE, message=FALSE, warning=FALSE}

max = avg_5_min %>% top_n(1, avg)

max
```


**6. Code to describe and show a strategy for imputing missing data**

When evaluating data, it is possible to see that column "steps" from original data is missing 2304 observations. The imputation method will be taking 5-min interval average to fill all missing values, as it can be seen in the next code chunk. After imputing values, the new dataset does not contain "NA's" anymore.

```{r, echo=TRUE, message=FALSE, warning=FALSE}

sapply(data, function(x) sum(is.na(x)))

data_na= data %>%
     group_by(interval) %>% 
      mutate_each(funs(replace(., which(is.na(.)),
                                mean(., na.rm=TRUE))))

sapply(data_na, function(x) sum(is.na(x)))

```
**7. Histogram of the total number of steps taken each day after missing values are imputed**

After imputing values, it is possible to see that distribution changed a little bit, more cases fall into the middle bin.

```{r, echo=TRUE, message=FALSE, warning=FALSE}

total_stepsb = data_na %>% group_by(date) %>% summarise(total=sum(steps))

ggplot(total_stepsb, aes(total)) + geom_histogram(bins=7, col="white", fill="navyblue", alpha=.8)+labs(title="Steps taken per day")

```

**Mean and median after imputing values with method described above**

Mean and median were not really affected after imputing values, actually only median changed from 10,765 to 10,766.19.
```{r, echo=TRUE, message=FALSE, warning=FALSE}

mean(total_stepsb$total)
median(total_stepsb$total)
```

**8. Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends**

Contrary to the pattern during weekdays, on weekends, more activity is registered in the afternoon. For weekdays, only the morning shows a higher level of activity.
```{r, echo=TRUE, message=FALSE, warning=FALSE}

data_wk = data %>% mutate(wk= as.numeric(isWeekday(date, wday=1:5)))

data_wk$wk=replace(data_wk$wk,data_wk$wk==1, "Weekdays")

data_wk$wk=replace(data_wk$wk,data_wk$wk==0, "Weekends")

avg_5_mine = data_wk  %>% na.omit %>% group_by(interval,wk) %>% summarise(avg=mean(steps))

ggplot(avg_5_mine, aes(x=interval,y=avg))+geom_line() + facet_grid(.~wk)

```

